When Subgraph Isomorphism is Really Hard, and Why This Matters for Graph Databases

### 2017-11-10 paper/paper.tex
Some re-wording

We therefore believe it would be unlikely to cause too much astonishment if we suggested that a
better subgraph isomorphism algorithm could be dropped in as a black box replacement in graph
databases systems to improve their performance.  This is not our claim. Instead, this paper shows
that better algorithms invalidate the flawed premise underlying the entire filter~/ verify approach.
We will now demonstrate empirically that filter~/ verify is simply a poor workaround for the kinds
of deficiency in VF2 demonstrated towards the end of the previous section.

@@

There are at least four benefits beyond simplicity towards abolishing filter~/ verify in favour of a
purely constraint programming inspired approach:

@@

Unfortunately, indexing is not the only incorrect design choice being made in systems built around
subgraph isomorphism algorithms, and lessons from constraint programming can be applied to other
subgraph-related research. We illustrate this by looking at several recent papers on subgraph
isomorphism, discussing how a deeper understanding of the empirical hardness of subgraph isomorphism
could lead the reader to different conclusions than the ones presented.

\citet{DBLP:conf/edbt/KatsarouNT17} look at what they call ``straggler'' queries, which they define
to be the small subset of queries that they observe taking much longer than others to solve. The
authors present an approach to address this perceived problem that they describe as novel: ``instead
of trying to come up with new algorithms for sub-iso testing, we utilize isomorphic query rewritings
and existing alternative algorithms in parallel''. Their first claim is that permuting graphs before
running the subgraph isomorphism algorithm can lead to ``wildly different'' execution times. They
note that VF2 does not define ``a strict order in which the nodes of the query are matched'', and so
permuting the graphs (for example, by using degree or label frequency information) before search can
sometimes give orders of magnitude improvements. The connection to variable- and value-ordering
heuristics is not noted, and no consideration is given to even the simplest dynamic ordering
heuristics like ``smallest domain first'' \citep{DBLP:journals/ai/HaralickE80}. A little thought
shows that of the orderings proposed, those involving placing rare labels first are effectively poor
approximations to a static ``smallest domain at top of search first'' ordering, whilst the remainder
use degree as we did earlier in this paper. However, \citeauthor{DBLP:conf/edbt/KatsarouNT17} do not
investigate any algorithm which employs domains, let alone strong variable- or value-ordering
heuristics or all-different propagation, and do not consider that their apparent successes could be
due to VF2's weaknesses rather than an inherent property of \NP-completeness.

@@

\citet{DBLP:conf/edbt/KatsarouNT17}'s second claim is that ``different
algorithms find different queries hard''. To address this, they run many subgraph isomorphism
algorithms and input permutations in parallel; the extensive literature on parallel portfolios in
general \citep{DBLP:journals/ai/GomesS01}, and the approach by \citet{DBLP:conf/sls/BattitiM07} for
subgraph isomorphism in particular, is not noted (and nor are portfolios mentioned when they say
that ``using machine learning models to predict which version \ldots to employ per query is of high
interest'').  The evidence so far in this paper suggests that we should be wary. It is certainly
true that there are some instances that all algorithms find hard, and there are good theoretical
reasons to believe that these instances genuinely are really hard.  Furthermore, algorithm
portfolios are also well-known to be a successful technique, and are beneficial even with modern
subgraph isomorphism algorithms \citep{DBLP:conf/lion/KotthoffMS16}.  However, in
\cref{section:labelled} we saw that there are many instances that VF2 finds hard that should not be
hard, and that are not hard for other algorithms.  If permuting graphs sometimes addresses the
difficulty of some of these instances, then it is likely that they are not genuinely hard instances
at all, and perhaps it would be better simply to start using an algorithm with domain tracking and
domain-based ordering heuristics, rather than assembling a portfolio of bad algorithms in the hopes
that at least one of them will often get lucky.

@@

Unfortunately, this lesson has not sunk in. For example, \citet{DBLP:journals/tcbb/BonniciG17}
recently compared various non-constraint-based approaches to LAD using benchmark sets whose
difficulties are measured between microseconds and milliseconds. Their results suggest that LAD is
an order of magnitude slower than other approaches. This should not surprise us: the initialisation
costs of LAD are non-trivial due to it constructing domains, calculating neighbourhood degree
sequences, and performing all-different filtering. However, the risks of occasionally encountering
exponential performance from weaker algorithms on easy instances should now be clear.

@@

Unfortunately, we saw that this aspect of VF2's performance has had practical consequences, most
notably being the misdesign of graph database systems. By not trying even the simplest constraint
programming techniques from the literature, and ignoring all that is known about the empirical
hardness of \NP-complete problems, members of the graph databases community have misled themselves
into believing that extensive research into supporting techniques is important, whereas really all
they are doing is working around some but not all of the defects in their choices of subgraph
isomorphism algorithms. With this new understanding of what does and does not make subgraph
isomorphism hard, it is time for a radical rethink of how graph database systems work.

@@

We must stress that we are not simply concluding that graph database systems should use a better
subgraph isomorphism algorithm.  Instead, this paper has shown that such systems need to be designed
respecting our understanding of the empirical hardness of \NP-complete problems, that subgraph
isomorphism algorithms should not be treated as a black box, and that lessons learned in constraint
programming and artificial intelligence should not go unheeded in other domains.

>>>>

We therefore believe it would be unlikely to cause too much astonishment if we suggested that a
better subgraph isomorphism algorithm could be dropped in as a black box replacement in graph
databases systems to improve their performance.  This is not our claim. Instead, this paper shows
that better algorithms invalidate the premise underlying the entire filter~/ verify approach.
We will now demonstrate empirically that filter~/ verify is simply a poor workaround for the kinds
of deficiency in VF2 demonstrated towards the end of the previous section.

@@

There are at least four benefits beyond simplicity towards abandoning filter~/ verify in favour of a
purely constraint programming inspired approach:

@@

Indexing is not the only incorrect design choice being made in systems built around
subgraph isomorphism algorithms, and lessons from constraint programming can be applied to other
subgraph-related research. We illustrate this by looking at several recent papers on subgraph
isomorphism, discussing how a deeper understanding of the empirical hardness of subgraph isomorphism
could lead the reader to different conclusions than the ones presented.

\citet{DBLP:conf/edbt/KatsarouNT17} look at ``straggler'' queries, which they define
to be the small subset of queries that they observe taking much longer than others to solve. The
authors present an approach to address this perceived problem that they describe as novel: ``instead
of trying to come up with new algorithms for sub-iso testing, we utilize isomorphic query rewritings
and existing alternative algorithms in parallel''. Their first claim is that permuting graphs before
running the subgraph isomorphism algorithm can lead to ``wildly different'' execution times. They
note that VF2 does not define ``a strict order in which the nodes of the query are matched'', and so
permuting the graphs (for example, by using degree or label frequency information) before search can
sometimes give orders of magnitude improvements. The connection to variable- and value-ordering
heuristics is not noted, and no consideration is given to dynamic ordering
heuristics like ``smallest domain first'' \citep{DBLP:journals/ai/HaralickE80}.
Of the orderings proposed, those involving placing rare labels first are effectively poor
approximations to a static ``smallest domain at top of search first'' ordering, whilst the remainder
use degree as we did earlier in this paper. However, \citeauthor{DBLP:conf/edbt/KatsarouNT17} do not
investigate any algorithm which employs domains, let alone strong variable- or value-ordering
heuristics or all-different propagation, and do not consider that their apparent successes could be
due to VF2's weaknesses rather than an inherent property of \NP-completeness.

@@

\citet{DBLP:conf/edbt/KatsarouNT17}'s second claim is that ``different
algorithms find different queries hard''. To address this, they run many subgraph isomorphism
algorithms and input permutations in parallel; the extensive literature on parallel portfolios in
general \citep{DBLP:journals/ai/GomesS01}, and the approach by \citet{DBLP:conf/sls/BattitiM07} for
subgraph isomorphism in particular, is not noted (and nor are portfolios mentioned when they say
that ``using machine learning models to predict which version \ldots to employ per query is of high
interest'').  The evidence so far in this paper suggests that we should be wary. It is certainly
true that there are some instances that all algorithms find hard, and there are good theoretical
reasons to believe that these instances genuinely are really hard.  Furthermore, algorithm
portfolios are also well-known to be a successful technique, and are beneficial even with modern
subgraph isomorphism algorithms \citep{DBLP:conf/lion/KotthoffMS16}.  However, in
\cref{section:labelled} we saw that there are many instances that VF2 finds hard that should not be
hard, and that are not hard for other algorithms.  If permuting graphs sometimes addresses the
difficulty of some of these instances, then it is likely that they are not genuinely hard instances
at all, and perhaps it would be better simply to start using an algorithm with domain tracking and
domain-based ordering heuristics, rather than assembling a portfolio of weak algorithms in the hopes
that at least one of them will often get lucky.

@@

Unfortunately, this lesson has not been well heeded. For example, \citet{DBLP:journals/tcbb/BonniciG17}
recently compared various non-constraint-based approaches to LAD using benchmark sets whose
difficulties are measured between microseconds and milliseconds. Their results suggest that LAD is
an order of magnitude slower than other approaches. This should not surprise us: the initialisation
costs of LAD are non-trivial due to it constructing domains, calculating neighbourhood degree
sequences, and performing all-different filtering. However, the risks of occasionally encountering
exponential performance from weaker algorithms on easy instances should now be clear.

@@

Unfortunately, we saw that this aspect of VF2's performance has had practical consequences, most
notably being the misdesign of graph database systems. By not trying even the simplest constraint
programming techniques from the literature, and ignoring all that is known about the empirical
hardness of \NP-complete problems, members of the graph databases community have reached the mistaken
conclusion that extensive research into supporting techniques is important, whereas really all
they are doing is working around some but not all of the defects in their choices of subgraph
isomorphism algorithms. With this new understanding of what does and does not make subgraph
isomorphism hard, it is time for a radical rethink of how graph database systems work.

@@

We must stress that we are not simply concluding that graph database systems should use a better
subgraph isomorphism algorithm.  Instead, this paper has shown that such systems need to be designed
in light of a modern understanding of the empirical hardness of \NP-complete problems, that subgraph
isomorphism algorithms should not be treated as a black box, and that lessons learned in constraint
programming and artificial intelligence should not go unheeded in other domains.
